---
title: "fish_encrusting_sponges_ecology"
author: "SGM"
date: '2025-04-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Attach the packages
```{r}
library(betareg)
library(grid)
library(dplyr)
library(gridExtra)
library(scales)
library(car)
library(ggplot2)
library(ggfortify)
library(broom)
library(effects)
library(emmeans)
library(MASS)
library(MuMIn)
library(devtools)
library(forcats)
library(tidyverse)
library(viridis)
library(nlme)
library(extrafont)
library(DHARMa)
library(ggrepel)
library(glmmTMB)
library(corrplot)
library(caret)
```

FISHES
##1. loading the data for all the species
```{r}
spofish <- read.csv("TO_BE_SPECIFIED_BY_USER/fish_asse_sp.csv", head=T, stringsAsFactors = F, fileEncoding = "latin1")
```

## 1 ENVIRONMENTAL VARIABLES
```{r}
env_nmds <- read.csv("TO_BE_SPECIFIED_BY_USER/environmental_data_all.csv", head=T, stringsAsFactors = F, fileEncoding = "latin1")
```

##1.3.Transforming variables Sampling and Treatment Into a factor
```{r}
spofish=spofish%>%mutate(Sampling=factor(Sampling),
                         Treatment=factor(Treatment, levels=c("Low", "Med", "High")))
```

2. EPLORATORY DATA ANALYSES
This will describe and summarize some metrics of the variables and exploratory plots to detect some trends in the data that could suggest how to follow up on this.
```{r}
spofishob <- spofish[spofish$Treatment%in% c("Low","Med", "High"), ]     
spofishex <- spofish[spofish$Treatment == "Removal", ] 
```

Sub-setting to separate observational and experimental data in different matrices for environmental variables
```{r}
env_ob_nmds<-env_nmds[env_nmds$Treatment %in% c("Low", "Med", "High"),]
env_ex_nmds<- env_nmds[env_nmds$Treatment=="Removal",]
```

#3 TOTAL ABUNDANCE AND SPECIES RICHNESS BY SPONGE COVER
```{r}
SE=function(x) {sd(x)/sqrt(length(x))}#standard error function
```

```{r}
spofishob$ID<-seq.int(nrow(spofishob))
```

```{r}
spofishobsum<-spofishob %>% rowwise (ID) 
```

```{r}
spofishobsum <-spofishobsum %>% mutate(total=sum(c_across(5:183)))# abundance
ab <- spofishobsum %>% group_by(Treatment) %>% summarize(Abundance=mean(total), SE=SE(total)) %>% ungroup()
```

```{r}
spofishobsum<-spofishobsum %>% mutate(Richness=sum(c_across(5:183)>0))# species richness
ri<-spofishobsum %>% group_by (Treatment) %>% summarize(Richnessmean=mean(Richness), SEr=SE(Richness))
```

8.1 ABUNDANCE MODEL:DATA MATRIX WITH ABUNDANCE, RICHNESS AND ALL ENVIRONMENTAL VARIABLES
```{r}
Abundance = spofishobsum$total# column of abundance 144 points
Richness = spofishobsum$Richness# column of Richness 144 points
Sampling = spofishobsum$Sampling# column of Sampling 144 points
Sponge_coverf = spofishobsum$Treatment# column of treatment 144 points
Quadrat = spofishobsum$Quadrat# column of Quadrat 144 points
Quadrat_code = spofishobsum$Quadrat_code# column of Quadrat_code 144 points
Depth = env_ob_nmds$Depth# column of depth 144 points
Sponge_cover = env_ob_nmds$Sponge_cover # column of sponge_cover 144 points
Rugosity = env_ob_nmds$ Rugosity# column of rugosity 144 points
Coral_cover = env_ob_nmds$ Coral_Cover # column of coral_cover 144 points
Other_cover = env_ob_nmds$ Other_cover # column of other_cover 144 points
biototal_env = data.frame (Abundance, Richness, Sampling, Sponge_coverf, Quadrat, Quadrat_code, Depth, Sponge_cover, Rugosity, Coral_cover, Other_cover)
```

```{r}
set.seed(8)
```

```{r}
#Correlations
enviro_cor <- biototal_env [,c(7:11)]
enviro_cor2 <- na.omit(enviro_cor)
c <-cor(enviro_cor2)
corrplot(c, method = 'number', order = 'alphabet')
cor_spon <- findCorrelation(
  c,
  cutoff = 0.6,
  verbose = TRUE,
  names = TRUE,
  exact = TRUE # Whether average correlation should be recomputed at each step
)
cor_spon
```

```{r}
biototal_env<-biototal_env %>% mutate(Sponge_coverf = factor(Sponge_coverf, levels=c("Low", "Med", "High"))) 
```

```{r}
sp_fish1f<-glmmTMB(Abundance~Sponge_coverf+Rugosity+Coral_cover+Depth+(1|Quadrat/Sampling), data=biototal_env,
                  family=nbinom2(link = "log"),  REML=TRUE)#nbinom2
```

```{r}
sp_fish1f_res<-simulateResiduals(sp_fish1f)
plot(sp_fish1f_res)
```

```{r}
Anova(sp_fish1f)
summary(sp_fish1f)
```

```{r}
plot(allEffects(sp_fish1f, type="response"))
```

```{r}
sp_fish1f_grid = with(biototal_env, list(Sponge_coverf = levels(Sponge_coverf), len=100))
```

```{r}
sp_fish1f.data = emmeans(sp_fish1f, ~ Sponge_coverf, at=sp_fish1f_grid, type='response') %>%  as.data.frame
```

```{r, warning=F}
sp_fish1f.sum<-sp_fish1f.data%>%
  group_by(Sponge_coverf) %>% 
  median_hdci(.width=c(0.8,0.95))
```

```{r, GB-fig3}
f1<-ggplot(sp_fish1f.data, aes(y=response, x=Sponge_coverf, color=Sponge_coverf))+
    geom_pointrange(aes(ymin=asymp.LCL, ymax=asymp.UCL), position=position_dodge(width=1), size=1)+
    geom_errorbar(aes(ymin=asymp.LCL, ymax=asymp.UCL), position=position_dodge(width=0.9), size=0.2, width=0.2)+
  scale_y_continuous(breaks=seq(0,250,50), limits=c(0,250))+
  scale_color_manual(values = c("darkorange2","#35978f","hotpink4")) +
  scale_fill_manual(values = c("darkorange2","#35978f","hotpink4"))+
  ylab("Abundance (Mean±95%CI)\n")+
  xlab("\nSponge cover")+
  labs(colour="Treatment")+
  theme(legend.position =c("none"), 
        axis.title.x = element_text(color="black", size=18, face=1),
        axis.title.y = element_text(color="black", size=18, face=1),
        axis.text.x = element_text(colour = "black", size=14, face=1), 
        axis.text.y = element_text(colour = "black", size=14, face=1 ),  
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f1 +
geom_point(data=biototal_env, aes(x = Sponge_coverf, y=Abundance, fill= Sponge_coverf, alpha= 0.5), size=3,   shape = 21, col="black", position = position_jitterdodge(jitter.width = 1.5, jitter.height=0.4, dodge.width=0.9))
```

```{r}
sp_fish1f_grid1 = with(biototal_env, list(Rugosity = seq(min(Rugosity), max(Rugosity), len=100)))
```

```{r}
sp_fish1f.data1 = emmeans(sp_fish1f, ~Rugosity, at=sp_fish1f_grid1, type='response') %>%
    as.data.frame
```

```{r}
f2<-ggplot(sp_fish1f.data1 , aes(x=Rugosity)) +
  geom_line(data= sp_fish1f.data1, aes(y=response), size=1, color="black", fill="black") +
  geom_ribbon(data=sp_fish1f.data1,  aes(ymin=asymp.LCL, ymax=asymp.UCL), color="#35978f", fill="#35978f", alpha=0.6)+
  scale_y_continuous(breaks=seq(0,250,50), limits=c(0,250))+
  scale_x_continuous(breaks=seq(0,1,0.1), limits=c(0.4,0.8))+
  ylab("Abundance (Mean ± 95% CI)\n")+
  xlab("\nSubstratum complexity")+
  theme(legend.position =c(), 
        axis.title.x = element_text(color="black", size=18, face="bold"),
        axis.title.y = element_text(color="black", size=18, face="bold"),
        axis.text.x = element_text(colour = "black", size=14, face="bold"),
        axis.text.y = element_text(colour = "black", size=14, face="bold" ), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f2+geom_point(data=biototal_env, aes(x = Rugosity, y=Abundance, fill= Sponge_coverf), size=3,   shape = 21, col="black")+
  scale_color_manual(values = c("darkorange2","#35978f","hotpink4")) +
  scale_fill_manual(values = c("darkorange2","#35978f","hotpink4"))+
  labs(fill="Sponge cover", colour="Sponge cover")#+
```

```{r}
sp_fish1f_grid1 = with(biototal_env, list(Coral_cover = seq(min(Coral_cover), max(Coral_cover), len=100)))
```

```{r}
sp_fish1f.data1 = emmeans(sp_fish1f, ~ Coral_cover, at=sp_fish1f_grid1, type='response') %>%
    as.data.frame
```

```{r}
f3 <-ggplot(sp_fish1f.data1, aes(x = Coral_cover)) +
  geom_line(data= sp_fish1f.data1, aes(y=response), size=1, color="black", fill="black") +
  geom_ribbon(data=sp_fish1f.data1,  aes(ymin=asymp.LCL, ymax=asymp.UCL), color="#35978f", fill="#35978f", alpha=0.6)+
  scale_y_continuous(breaks=seq(0,250,50), limits=c(0,250))+
  scale_x_continuous(breaks=seq(0,100,10), limits=c(10,90))+
  ylab("Abundance (Mean ± 95% CI)\n")+
  xlab("\nCoral Cover (%)")+
  theme(legend.position =c(), 
        axis.title.x = element_text(color="black", size=18, face="bold"),
        axis.title.y = element_text(color="black", size=18, face="bold"),
        axis.text.x = element_text(colour = "black", size=14, face="bold"),
        axis.text.y = element_text(colour = "black", size=14, face="bold" ), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f3+geom_point(data=biototal_env, aes(x = Coral_cover, y=Abundance, fill= Coral_cover), size=3,   shape = 21, col="black")+
  scale_fill_distiller(palette = "YlOrBr", trans = "reverse")
```

```{r}
sp_fish1f_grid1 = with(biototal_env, list(Depth = seq(min(Depth), max(Depth), len=100)))
```

```{r}
sp_fish1f.data1 = emmeans(sp_fish1f, ~ Depth, at=sp_fish1f_grid1, type='response') %>%
    as.data.frame
```

```{r}
f4 <-ggplot(sp_fish1f.data1, aes(x = Depth)) +
  geom_line(data= sp_fish1f.data1, aes(y=response), size=1, color="black", fill="black") +
  geom_ribbon(data=sp_fish1f.data1,  aes(ymin=asymp.LCL, ymax=asymp.UCL), color="#35978f", fill="#35978f", alpha=0.6)+
  scale_y_continuous(breaks=seq(0,250,50), limits=c(0,250))+
  scale_x_continuous(breaks=seq(5,10,1), limits=c(5,10))+
  ylab("Abundance (Mean ± 95% CI)\n")+
  xlab("\nDepth (m)")+
  theme(legend.position =c(), 
        axis.title.x = element_text(color="black", size=18, face="bold"),
        axis.title.y = element_text(color="black", size=18, face="bold"),
        axis.text.x = element_text(colour = "black", size=14, face="bold"),
        axis.text.y = element_text(colour = "black", size=14, face="bold" ), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f4+geom_point(data=biototal_env, aes(x = Depth, y=Abundance, fill= Depth), size=3,   shape = 21, col="black")+
  scale_fill_distiller(trans = "reverse")
```

# 8.2 SPECIES RICHNESS MODEL:DATA MATRIX WITH ABUNDANCE, RICHNESS AND ALL ENVIRONMENTAL VARIABLES
```{r}
spri_fish1f<-glmmTMB(Richness~Sponge_coverf+Rugosity+Coral_cover+Depth+(1|Quadrat/Sampling), data=biototal_env,
                  family=nbinom2(link = "log"),  REML=TRUE)
```

```{r}
spri_fish1f_res<-simulateResiduals(sp_fish1f)
plot(spri_fish1f_res)
```

```{r}
Anova(spri_fish1f)
summary(spri_fish1f)
```

```{r}
plot(allEffects(spri_fish1f, type="response"))
```

```{r}
spri_fish1f_grid = with(biototal_env, list(Sponge_coverf = levels(Sponge_coverf), len=100))
```

```{r}
spri_fish1f.data = emmeans(spri_fish1f, ~Sponge_coverf, at=spri_fish1f_grid, type='response') %>%
    as.data.frame
spri_fish1f.data 
```

```{r, GB-fig3}
f5<-ggplot(spri_fish1f.data, aes(y=response, x=Sponge_coverf, color=Sponge_coverf))+
    geom_pointrange(aes(ymin=asymp.LCL, ymax=asymp.UCL), position=position_dodge(width=0.9), size=1)+
    geom_errorbar(aes(ymin=asymp.LCL, ymax=asymp.UCL), position=position_dodge(width=0.9), size=0.1, width=0.1)+
  scale_y_continuous(breaks=seq(0,40,10), limits=c(0,40))+
  scale_color_manual(values = c("darkorange2","#35978f","hotpink4")) +
  scale_fill_manual(values = c("darkorange2","#35978f","hotpink4"))+
  ylab("Species richness (Mean ± 95% CI)\n")+
  xlab("\nSponge cover")+
  labs(colour="Treatment")+
  theme(legend.position =c("none"), 
        axis.title.x = element_text(color="black", size=21, face=1),
        axis.title.y = element_text(color="black", size=21, face=1),
        axis.text.x = element_text(colour = "black", size=16, face=1),
        axis.text.y = element_text(colour = "black", size=16, face=1), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f5 + geom_point(data=biototal_env, aes(x = Sponge_coverf, y=Richness, fill= Sponge_coverf, alpha= 0.5), size=3,   shape = 21, col="black", position = position_jitterdodge(jitter.width = 1.5, jitter.height=0.4, dodge.width=0.9))
```

```{r}
spri_fish1f_grid1 = with(biototal_env, list(Rugosity = seq(min(Rugosity), max(Rugosity), len=100)))
```

```{r}
spri_fish1f.data1 = emmeans(spri_fish1f, ~ Rugosity, at = spri_fish1f_grid1, type='response') %>%as.data.frame
```

```{r}
f6<-ggplot(spri_fish1f.data1 , aes(x=Rugosity)) +
  geom_line(data= spri_fish1f.data1, aes(y=response), size=1, color="black", fill="black") + geom_ribbon(data=spri_fish1f.data1,  aes(ymin=asymp.LCL, ymax=asymp.UCL), color="#35978f", fill="#35978f", alpha=0.6)+
  scale_y_continuous(breaks=seq(0,40,10), limits=c(0,40))+
  scale_x_continuous(breaks=seq(0,1,0.1), limits=c(0.4,0.8))+
  ylab("Species richness (Mean±95%CI)\n")+
  xlab("\nSubstratum rugosity")+
    theme(legend.position =, 
        axis.title.x = element_text(color="black", size=18, face=1),
        axis.title.y = element_text(color="black", size=18, face=1),
        axis.text.x = element_text(colour = "black", size=14, face=1),
        axis.text.y = element_text(colour = "black", size=14, face=1), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face=1, size=14))
f6+geom_point(data=biototal_env, aes(x = Rugosity, y=Richness, fill= Sponge_coverf), size=3,   shape = 21,col="black")+
  scale_color_manual(values = c("darkorange2","#35978f","hotpink4")) +
  scale_fill_manual(values = c("darkorange2","#35978f","hotpink4"))+
  labs(fill="Sponge cover", colour="Sponge cover")#+
```

```{r}
spri_fish1f_grid = with(biototal_env, list(Coral_cover = seq(min(Coral_cover), max(Coral_cover), len=100)))
```

```{r}
spri_fish1f.data = emmeans(spri_fish1f, ~ Coral_cover, at=spri_fish1f_grid, type='response') %>%
    as.data.frame
```

```{r}
f7 <-ggplot(spri_fish1f.data, aes(x = Coral_cover)) +
  geom_line(data= spri_fish1f.data, aes(y=response), size=1, color="black", fill="black") +
  geom_ribbon(data=spri_fish1f.data,  aes(ymin=asymp.LCL, ymax=asymp.UCL), color="#35978f", fill="#35978f", alpha=0.6)+
  scale_y_continuous(breaks=seq(0,40,10), limits=c(0,40))+
  scale_x_continuous(breaks=seq(0,100,10), limits=c(10,90))+
  ylab("Species richness (Mean±95%CI)\n")+
  xlab("\nCoral Cover (%)")+
  theme(legend.position =c(), 
        axis.title.x = element_text(color="black", size=18, face="bold"),
        axis.title.y = element_text(color="black", size=18, face="bold"),
        axis.text.x = element_text(colour = "black", size=14, face="bold"),
        axis.text.y = element_text(colour = "black", size=14, face="bold" ), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f7+geom_point(data=biototal_env, aes(x = Coral_cover, y=Richness, fill= Coral_cover), size=3,   shape = 21, col="black")+
  scale_fill_distiller(palette = "YlOrBr", trans = "reverse")
```

```{r}
spri_fish1f_grid = with(biototal_env, list(Depth = seq(min(Depth), max(Depth), len=100)))
```

```{r}
spri_fish1f.data = emmeans(spri_fish1f, ~ Depth, at=spri_fish1f_grid, type='response') %>%
    as.data.frame
```

```{r}
f8 <-ggplot(spri_fish1f.data, aes(x = Depth)) +
  geom_line(data= spri_fish1f.data, aes(y=response), size=1, color="black", fill="black") +
  geom_ribbon(data=spri_fish1f.data,  aes(ymin=asymp.LCL, ymax=asymp.UCL), color="#35978f", fill="#35978f", alpha=0.6)+
  scale_y_continuous(breaks=seq(0,40,10), limits=c(0,40))+
  scale_x_continuous(breaks=seq(5,10,1), limits=c(5,10))+
  ylab("Species richness (Mean±95%CI)\n")+
  xlab("\nCoral Cover")+
  theme(legend.position =c(), 
        axis.title.x = element_text(color="black", size=18, face="bold"),
        axis.title.y = element_text(color="black", size=18, face="bold"),
        axis.text.x = element_text(colour = "black", size=14, face="bold"),
        axis.text.y = element_text(colour = "black", size=14, face="bold" ), 
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white", color = NA))+
        theme(text=element_text(family="Times", face="bold", size=12))
f8+geom_point(data=biototal_env, aes(x = Depth, y=Richness, fill= Depth), size=3,   shape = 21, col="black")+
  scale_fill_distiller(trans = "reverse")
```





