---
title: "fish_encrusting_sponges_ecology"
author: "SGM"
date: '2025-04-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Attach the packages
```{r}
library(betareg)
library(grid)
library(dplyr)
library(gridExtra)
library(scales)
library(car)
library(ggplot2)
library(ggfortify)
library(broom)
library(vegan)
library(devtools)
library(tidybayes)
library(forcats)
library(tidyverse)
library(BiodiversityR) 
library(nlme)
library(extrafont)
```

FISHES
##1. loading the data for all the species
```{r}
spofish <- read.csv("TO_BE_SET_BY_USER/fish_asse_sp.csv", head=T, stringsAsFactors = F, fileEncoding = "latin1")
```

## 1 ENVIRONMENTAL VARIABLES
```{r}
env_nmds <- read.csv("TO_BE_SET_BY_USER/environmental_data_all.csv", head=T, stringsAsFactors = F, fileEncoding = "latin1")
```

##1.3.Transforming variables Sampling and Treatment Into a factor
```{r}
spofish=spofish%>%mutate(Sampling=factor(Sampling),
                         Treatment=factor(Treatment, levels=c("Low", "Med", "High")))
```

2. DATA PREPARATION
```{r}
spofishob <- spofish[spofish$Treatment%in% c("Low","Med", "High"), ]     
spofishex <- spofish[spofish$Treatment == "Removal", ] 
```
Sub-setting to separate observational and experimental data in different matrices for environmental variables
```{r}
env_ob_nmds<-env_nmds[env_nmds$Treatment %in% c("Low", "Med", "High"),]
env_ex_nmds<- env_nmds[env_nmds$Treatment=="Removal",]
```

```{r}
sp_ob_nmdsr <- spofishob[,c(-1,-2,-3,-4,-5)]
env_ob_nmsr <- env_ob_nmds[,c(-1,-2,-3,-4,-5)]
```

#MULTIVARIATE ANALYSIS NMDS

### 2. TRANSFORMATION FOR THE DATA
```{r}
spofishob.TRAN.WIS=wisconsin(spofishob[,c(-1,-2,-3,-4)])
spofishobfactor=spofishob[,-c(3:144)]
```

### 3 MEASURES OF DISTANCE: BRAY CURTIS
```{r}
dist.WISob.bray<- vegdist(spofishob.TRAN.WIS, method="bray")
```

#######################################################################################################################################

###  ADONIS Permutational Multivariate Analysis of Variance - adonis Fish on sponges 
###  Environmental data preparation 

Data matrix only with my variables of interest Sampling
```{r}
env_ob_nmds2 =(env_ob_nmds[,c(-4:-10)])
env_ob_nmds2<-env_ob_nmds2 %>% mutate(Treatment=factor(Treatment))
env_ob_nmds <- env_ob_nmds %>% mutate(Treatment=factor(Treatment),
                                      Sampling=factor(Sampling))
```

```{r}
h1 <- with(env_ob_nmds, how(nperm = 999, blocks = Sampling))
env_ob_nmds.adonis <- adonis2(dist.WISob.bray~ Treatment+Rugosity+ Coral_Cover+ Depth, data=env_ob_nmds, permutations = h1)
env_ob_nmds.adonis 
```
#######################################################################################################################################

## RDA 1 FISH COMMUNITY AND SPONGE COVER AND STRUCTURAL VARIABLES
```{r}
sp_ob_rda <- sp_ob_nmdsr
env_ob_rda <- env_ob_nmds %>% mutate(Treatment=factor(Treatment))
```

```{r}
attach(env_ob_rda)
```

```{r}
f_hel <- disttransform(sp_ob_rda, method='hellinger')# The function is in BiodiversityR not in vegan
```

```{r}
h1 <- with(env_ob_rda, how(nperm = 999, blocks = Sampling))
f_rda<- rda(f_hel ~Sponge_cover+ Rugosity + Coral_Cover+ Depth, data=env_ob_rda, permutations = h1) #scale = T,
```

```{r}
summary(f_rda)
```

```{r}
fwd.sel <- ordiR2step(rda(f_hel ~ 1, data = env_ob_rda), # lower model limit (simple!)
                      scope = formula(f_rda), # upper model limit (the "full" model)
                      direction = "forward",
                      R2scope = TRUE, # can't surpass the "full" model's R2
                      pstep = 1000,
                      trace = FALSE) # change to TRUE to see the selection process!
```

```{r}
fwd.sel$call
```

```{r}
RsquareAdj(f_rda)
```

```{r}
anova.cca(f_rda, step = 1000)
anova.cca(f_rda, step = 1000, by = "term")
anova.cca(f_rda, step = 1000, by = "axis")
```

```{r}
ordiplot(f_rda, scaling = 1)#type = "text"
```
#Scaling 1 shows similarities between objects in the response matrix.
#Sites (numbers) that are closer together have more similar communities.
#Species that are closer together occupy more sites in common.
```{r}
ordiplot(f_rda, scaling = 2)#type = "text"
```
#Scaling 2 shows the effects of explanatory variables.
#Longer arrows mean this variable strongly drives the variation in the community matrix.

```{r}
perc <- round(100*(summary(f_rda)$cont$importance[2, 1:2]), 2)
sc_si <- scores(f_rda, display="sites", choices=c(1,2), scaling=1)
sc_sp <- scores(f_rda, display="species", choices=c(1,2), scaling=1)
sc_bp <- scores(f_rda, display="bp", choices=c(1, 2), scaling=1)
```

```{r}
sc_si_df<- as.data.frame(sc_si)
sc_si_df$row_names<-row.names(sc_si_df)
colnames(sc_si_df)[3] = "site_names"
```

```{r}
sc_si_df_1 = data.frame(RDA1 = sc_si_df$RDA1, 
                        RDA2 = sc_si_df$RDA2, 
                        Treatment = env_ob_rda$Treatment,
                        Sponge_cover = env_ob_rda$Sponge_cover,
                        Rugosity = env_ob_rda$Rugosity, 
                        Coral_Cover = env_ob_rda$Coral_Cover,
                        Depth = env_ob_rda$Depth)
```

```{r}
sc_si_df_1 <- sc_si_df_1 %>% mutate (Treatment=factor(Treatment, levels=c("Low", "Med", "High")))
```

```{r}
sc_bp_df<- as.data.frame(sc_bp)
sc_bp_df$row_names<-row.names(sc_bp_df)
colnames(sc_bp_df)[3] = "Enviro_var"
```

```{r}
sc_sp_df<- as.data.frame(sc_sp)
sc_sp_df$row_names<-row.names(sc_sp_df)
```

```{r}
sc_sp_df_1=data.frame(RDA1 = sc_sp_df$RDA1, 
                      RDA2 = sc_sp_df$RDA2, 
                      Fish_sp = sc_sp_df$row_names)
```

```{r, warning=FALSE}
species_importance <- ordiplot(f_rda, choices=c(1,2))
spec.envfit <- envfit(species_importance, env=f_hel)
spec.data.envfit <- data.frame(r=spec.envfit$vectors$r, p=spec.envfit$vectors$pvals)
species.long2 <- species.long(species_importance, spec.data=spec.data.envfit)
species_importance_f <- species.long2[species.long2$p <= 0.05,  ]#
species_importance_f
```

```{r}
#write_xlsx(species_importance_f,"TO_BE_SET_BY_USER/species_rda_ob_p005_2nd.xlsx")
```

```{r}
BioR.theme <- theme(
  panel.background = element_blank(),
  panel.border =  element_blank(),
  panel.grid = element_blank(),
  axis.line = element_line("gray25"),
  text = element_text(size = 12),
  axis.text = element_text(size = 10, colour = "gray25"),
  axis.title = element_text(size = 14, colour = "gray25"),
  legend.title = element_text(size = 14),
  legend.text = element_text(size = 14),
  legend.key = element_blank())
```

```{r}
#pdf(file="C:/Users/sgonzalezmur/Desktop/Manusrevision/Fish_sponges_assemblage_ecology/Ecosphere_submission/Revisions/Figure_4_r001_p005.pdf", width=10, height=7, colormodel='cmyk')
all_plot <- ggplot() + 
  geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
  geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
  xlab("RDA1 69%") +
  ylab("RDA2 15%") +  
  scale_x_continuous(breaks=seq(-0.2,0.2,0.1), limits=c(-0.25,0.25), sec.axis = dup_axis(labels=NULL, name=NULL)) +
  scale_y_continuous(breaks=seq(-0.2,0.2,0.1), limits=c(-0.25,0.25), sec.axis = dup_axis(labels=NULL, name=NULL)) +
   geom_segment(data=sc_bp_df,
               aes(x=0, y=0, xend=(RDA1), yend=(RDA2)),
               colour="black", linewidth=0.3, arrow=arrow()) +
   geom_text_repel(data=sc_bp_df,
                  aes(x=(RDA1*1.3), y=(RDA2*1.3), label=Enviro_var),
                  colour="black") +

  geom_segment(data=species_importance_f,
               aes(x=0, y=0, xend=(axis1/1.5), yend=(axis2/1.5)),
               colour="green4", linewidth=0.3, arrow=arrow())+
  geom_text_repel(data=species_importance_f,
                  aes(x=(axis1/1.3), y=(axis2/1.3), label=species_importance_f$labels),
                  colour="black", size = 1)+
  
  # geom_point(data=species_importance_f,
  # aes(x=axis1/2, y=axis2/2),
  # color="black",
  # shape = 23,
  # fill = "green4",
  # size = 3) +
  # geom_text_repel(data = species_importance_f,
  #             aes(x=axis1/2, y=axis2/2, label = species_importance_f$labels),
  #            colour="black") +
  # 
  geom_point(data=sc_si_df_1,
             aes(x=RDA1, y=RDA2,
                 colour=sc_si_df_1$Treatment,
                 fill=sc_si_df_1$Treatment,
                 shape=sc_si_df_1$Treatment),
             size = 2) +

  scale_colour_manual (name= "Treatment",
                      labels=c("Low","Med","High"),
                      values=c("black", "black","black"))+ 
  scale_fill_manual   (name= "Treatment",
                      labels=c("Low","Med","High"),
                      values=c("darkorange2","#35978f","hotpink4"))+ 
  scale_shape_manual  (name = "Treatment",
                      labels=c("Low","Med","High"),
                      values = c(21,22,24))+
 
   stat_ellipse(data = sc_si_df_1, aes(x=RDA1, y= RDA2, fill = Treatment, color = Treatment), geom="polygon",level=0.95, alpha=0.2) +
    BioR.theme
all_plot
```

```{r}
all_plot2 <- ggplot() + 
  geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
  geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
  xlab("RDA1 69%") +
  ylab("RDA2 15%") +  
  scale_x_continuous(breaks=seq(-0.2,0.2,0.1), limits=c(-0.25,0.25), sec.axis = dup_axis(labels=NULL, name=NULL)) +
  scale_y_continuous(breaks=seq(-0.2,0.2,0.1), limits=c(-0.25,0.25), sec.axis = dup_axis(labels=NULL, name=NULL)) +
   geom_segment(data=sc_bp_df, 
               aes(x=0, y=0, xend=(RDA1), yend=(RDA2)), 
               colour="black", linewidth=0.3, arrow=arrow()) +
   geom_text_repel(data=sc_bp_df, 
                  aes(x=(RDA1*1.3), y=(RDA2*1.3), label=Enviro_var),
                  colour="black") +
  
   
  geom_segment(data=species_importance_f, 
               aes(x=0, y=0, xend=(axis1), yend=(axis2)), 
               colour="green4", linewidth=0.3, arrow=arrow())+
  geom_text_repel(data=species_importance_f, 
                  aes(x=(axis1*1.3), y=(axis2*1.3), label=species_importance_f$labels),
                  colour="black") +
  
  #geom_point(data=species_importance_f, 
             #aes(x=axis1/2, y=axis2/2),
             #color="black", 
             #shape = 23, 
             #fill = "green4",
             #size = 3) +
    #geom_text_repel(data = species_importance_f, 
     #             aes(x=axis1/2, y=axis2/2, label = species_importance_f$labels),
      #            colour="black") +
  
 scale_colour_manual (name= "Treatment",
                      labels=c("Low","Med","High"),
                      values=c("black", "black","black"))+ 
  scale_fill_manual   (name= "Treatment",
                      labels=c("Low","Med","High"),
                      values=c("darkorange2","#35978f","hotpink4"))+ 
  scale_shape_manual  (name = "Treatment",
                      labels=c("Low","Med","High"),
                      values = c(21,22,24))+
 
   stat_ellipse(data = sc_si_df_1, aes(x=RDA1, y= RDA2, fill = Treatment, color = Treatment), geom="polygon",level=0.95, alpha=0.2) +
    BioR.theme

all_plot2
```
 
