---
title: "fish_encrusting_sponges_ecology_ex_nmds_rda"
author: "SGM"
date: '2025-04-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Attach the packages
```{r, warning=FALSE}
library(betareg)
library(grid)
library(dplyr)
library(gridExtra)
library(scales)
library(car)
library(ggplot2)
library(ggfortify)
library(broom)
library(vegan)
library(devtools)
library(tidybayes)
library(forcats)
library(tidyverse)
library(BiodiversityR) 
library(nlme)
library(ggsci)
library(ggrepel)
library(ggforce)
```

##1. loading the data for all the species
```{r}
sp_initial_ex <- read.csv("TO_BE_SET_BY_USER/fish_asse_sp_cca_ex_sampling_1_4.csv", head=T, stringsAsFactors = F, fileEncoding = "latin1")
```

## 1 ENVIRONMENTAL VARIABLES
```{r}
env_initial_ex <- read.csv("TO_BE_SET_BY_USER/environmental_data_all_ex_sampling_1_4.csv", head=T, stringsAsFactors = F, fileEncoding = "latin1")
```

##2. DATA PREPARATION
```{r}
sp_ex_final<-sp_initial_ex#Species variables
env_ex_final<-env_initial_ex %>% mutate(Time = factor(Time, levels = c("B1", "A1", "A2", "A3")),
                                        Quadrat = factor(Quadrat),
                                        Treatment = factor (Treatment, levels = c("High", "Removal")),
                                        Sampling = factor(Sampling, levels = c("1", "2", "3", "4")))
```

```{r}
sp_ex_nmdsr <- sp_ex_final[,-c(1:7)]
env_ex_nmsr <- env_ex_final[,-c(1:7)]
```
To plot the species I will use the envfit function. In the process I will need to bring the df with the data that have been transformed (Wisonsin) Transformation has been applied. The second element is the nmds that has been already run on the data. 

#######################################################################################################################################
#MULTIVARIATE ANALYSIS NMDS
### 2. TRANSFORMATION FOR THE DATA
```{r}
spofishex.TRAN.WIS=wisconsin(sp_ex_nmdsr)
```

### 3. MEASURES OF DISTANCE: BRAY CURTIS
```{r}
dist.WISex.bray<- vegdist(spofishex.TRAN.WIS, method="bray")
```

### 4. nMDS WISCONSIN TRANSFORMATION AND BRAY CURTIS DISTANCE
```{r}
nmdsWISex<-metaMDS(dist.WISex.bray, k=2, maxit = 999, model=c("global"))
```

## 5. PLOT DIAGNOSTICS nMDS SQRT TRANSFORMATION AND BRAY CURTIS DISTANCE 
```{r}
 plot(nmdsWISex)
 nmdsWISex$stress
 goodness(nmdsWISex)
 stressplot(nmdsWISex)
```

### 6. NEW MATRIX
```{r}
MDS1exW = nmdsWISex$points[,1]
MDS2exW = nmdsWISex$points[,2]
NMDSexW = data.frame(NMDS1 = MDS1exW, NMDS2 = MDS2exW, Sampling= sp_ex_final$Sampling, Treatment = sp_ex_final$Treatment, Time= sp_ex_final$Time, Condition=sp_ex_final$Gtime)
```

```{r}
NMDSexW$inter <- paste(NMDSexW$Treatment, NMDSexW$Condition)
NMDSexW$inter2 <- paste(NMDSexW$Treatment, NMDSexW$Time)
```

```{r}
NMDSexW<-NMDSexW %>% mutate(Sampling=factor(Sampling), 
                            inter=factor(inter, levels=c("High Before", "High After", "Removal Before", "Removal After")), 
                            inter2=factor(inter2, levels=c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3")))
```

```{r, echo=FALSE, warning=FALSE}
spex.envfit <- envfit(nmdsWISex, spofishex.TRAN.WIS, permutations = 999) # this fits environmental vectors
```

###  8.ADONIS Permutational Multivariate Analysis of Variance - adonis Fish on sponges
```{r}
h1ex <- with(env_ex_final, how(nperm = 999, blocks = Sampling))
Eenv_ex_nmds.adonis_sup <- adonis2(dist.WISex.bray~ Treatment*Time+ Rugosity+ Coral_Cover+ Depth, data=env_ex_final, permutations = h1ex)
Eenv_ex_nmds.adonis_sup
```
#######################################################################################################################################
## RDA 1 FISH COMMUNITY AND SPONGE COVER AND STRUCTURAL VARIABLES ALL SPECIES
```{r}
sp_ex_final_r<-sp_initial_ex[,-c(1:7)]#Species variables
env_ex_final_r<-env_initial_ex# environmental variables
```

```{r}
attach(env_ex_final_r)
```

```{r}
f_hel_ex<- disttransform(sp_ex_final_r, method='hellinger')# The function is in BiodiversityR not in vegan
```

```{r}
h1ex <- with(env_ex_final, how(nperm = 999, blocks = Time))
f_rda_ex<- rda(f_hel_ex ~Treatment*Time+Sponge_cover+ Rugosity + Coral_Cover+ Depth, data=env_ex_final_r, permutations = h1ex) #scale = T,
```

```{r}
summary(f_rda_ex)
```

```{r}
fwd.sel_ex <- ordiR2step(rda(f_hel_ex ~ 1, data = env_ex_final_r), # lower model limit (simple!)
                      scope = formula(f_rda_ex), # upper model limit (the "full" model)
                      direction = "forward",
                      R2scope = TRUE, # can't surpass the "full" model's R2
                      pstep = 1000,
                      trace = FALSE) # change to TRUE to see the selection process!
```

```{r}
fwd.sel_ex$call
```

```{r}
RsquareAdj(f_rda_ex)
```

```{r}
anova.cca(f_rda_ex, step = 1000)
anova.cca(f_rda_ex, step = 1000, by = "term")
anova.cca(f_rda_ex, step = 1000, by = "axis")
```

```{r}
ordiplot(f_rda_ex, scaling = 1)#type = "text"
```
#Scaling 1 shows similarities between objects in the response matrix.
#Sites (numbers) that are closer together have more similar communities.
#Species that are closer together occupy more sites in common.
```{r}
ordiplot(f_rda_ex, scaling = 2, display = "cn")#type = "text"
centroids <-scores(f_rda_ex, display = "cn")
centroids
```
#Scaling 2 shows the effects of explanatory variables.
#Longer arrows mean this variable strongly drives the variation in the community matrix.

```{r}
perc <- round(100*(summary(f_rda_ex)$cont$importance[2, 1:2]), 2)
sc_si <- scores(f_rda_ex, display="sites", choices=c(1,2), scaling=1)
sc_sp <- scores(f_rda_ex, display="species", choices=c(1,2), scaling=1)
sc_bp <- scores(f_rda_ex, display="bp", choices=c(1, 2), scaling=1)
```

```{r}
sc_si_df<- as.data.frame(sc_si)
sc_si_df$row_names<-row.names(sc_si_df)
colnames(sc_si_df)[3] = "site_names"
```

```{r}
sc_si_df_1 = data.frame(RDA1 = sc_si_df$RDA1, 
                        RDA2 = sc_si_df$RDA2, 
                        Treatment = env_ex_final$Treatment,
                        Sponge_cover = env_ex_final$Sponge_cover,
                        Rugosity = env_ex_final$Rugosity, 
                        Coral_Cover = env_ex_final$Coral_Cover,
                        Depth = env_ex_final$Depth,
                        NMDSexW$inter,
                        NMDSexW$inter2)
```

```{r}
sc_si_df_1 <- sc_si_df_1 %>% mutate (Treatment=factor(Treatment, levels=c("High", "Removal")))
```

```{r}
sc_bp_df<- as.data.frame(sc_bp)
sc_bp_df$row_names<-row.names(sc_bp_df)
colnames(sc_bp_df)[3] = "Enviro_var"
```

```{r}
sc_sp_df<- as.data.frame(sc_sp)
sc_sp_df$row_names<-row.names(sc_sp_df)
```

```{r}
sc_sp_df_1=data.frame(RDA1 = sc_sp_df$RDA1, 
                      RDA2 = sc_sp_df$RDA2, 
                      Fish_sp = sc_sp_df$row_names)
```

```{r, warning=FALSE}
species_importance <- ordiplot(f_rda_ex, choices=c(1,2))
spec.envfit <- envfit(species_importance, env=f_hel_ex)
spec.data.envfit <- data.frame(r=spec.envfit$vectors$r, p=spec.envfit$vectors$pvals)
species.long2 <- species.long(species_importance, spec.data=spec.data.envfit)
species_importance_f <- species.long2[species.long2$p <= 0.05,  ]#& species.long2$r >= 0.1,
species_importance_f
```

```{r}
#write_xlsx(species_importance_f,"TO_BE_SET_BY_USER//species_rda_ex_005_001.xlsx")# DONE AND SAVED
```

```{r}
BioR.theme <- theme(
  panel.background = element_blank(),
  panel.border =  element_blank(),
  panel.grid = element_blank(),
  axis.line = element_line("gray25"),
  text = element_text(size = 12),
  axis.text = element_text(size = 10, colour = "gray25"),
  axis.title = element_text(size = 14, colour = "gray25"),
  legend.title = element_text(size = 14),
  legend.text = element_text(size = 14),
  legend.key = element_blank())
```

```{r}
all_plot_ex <- ggplot() + 
  geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
  geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
  xlab("RDA1 44%") +
  ylab("RDA2 16%") +  
  scale_x_continuous(breaks=seq(-0.3,0.3,0.1), limits=c(-0.3,0.3), sec.axis = dup_axis(labels=NULL, name=NULL),labels = scales::comma) +
  scale_y_continuous(breaks=seq(-0.3,0.3,0.1), limits=c(-0.3,0.3), sec.axis = dup_axis(labels=NULL, name=NULL), labels = scales::comma) +
  
  geom_segment(data=sc_bp_df,
               aes(x=0, y=0, xend=(RDA1), yend=(RDA2)),
               colour="black", linewidth=0.3, arrow=arrow()) +
   ggrepel::geom_text_repel(data=sc_bp_df,
                  aes(x=(RDA1*1.3), y=(RDA2*1.3), label=Enviro_var),
                  colour="black") +
  
   geom_segment(data=species_importance_f,
               aes(x=0, y=0, xend=(axis1), yend=(axis2)),
               colour="green4", linewidth=0.3, arrow=arrow())+
   ggrepel::geom_text_repel(data=species_importance_f,
                  aes(x=(axis1*1), y=(axis2*1), label=species_importance_f$labels),
                  colour="black", size=1) +

 # geom_point(data=centroids, 
 #             aes(x=RDA1/2, y=RDA2/2),
 #             color="black",
 #             shape = 23,
 #             fill = "yellow4",
 #             size = 3) +
 #  # ggrepel::geom_text_repel(data = species_importance_f, 
 #  #                 aes(x=axis1/2, y=axis2/2, label = species_importance_f$labels),
 #  #                colour="black") +
  
  geom_point(data=sc_si_df_1,
             aes(x=RDA1, y=RDA2,
                 colour=sc_si_df_1$NMDSexW.inter2,
                 fill=sc_si_df_1$NMDSexW.inter2,
                 shape=sc_si_df_1$NMDSexW.inter2),
                 size = 2) +

  scale_colour_manual(name= "Treatment and Sampling Time",
                      labels=c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3"),
                      values=c("black",  "black", "black",  "black", "black",  "black", "black",  "black")) +
  scale_fill_manual(name= "Treatment and Sampling Time",
                      labels=c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3"),
                      values=c("hotpink4","magenta2","orchid1", "deeppink2","hotpink4","gray47", "gainsboro","gray96")) +
  scale_shape_manual(name ="Treatment and Sampling Time",
                      labels = c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3"),
                      values = c(24,24,24,24,15,0,1,2))+

   #stat_ellipse(data = sc_si_df_1, aes(x=RDA1, y= RDA2, fill = sc_si_df_1$NMDSexW.inter2, color = sc_si_df_1$NMDSexW.inter2), geom="polygon",level=0.95, alpha=0.2) +
    #BioR.theme
 
  # stat_ellipse(data = sc_si_df_1, aes(x=RDA1, y= RDA2, fill = Treatment, color = Treatment), geom="polygon",level=0.95, alpha=0.2) +
  # scale_colour_manual(name= "Treatment and Sampling Time",
  #                     labels=c("High","Removal"),
  #                     values=c("black",  "black")) +
  # scale_fill_manual(name= "Treatment and Sampling Time",
  #                     labels=c("High", "Removal"),
  #                     values=c("hotpink4","gray47")) +
  # scale_shape_manual(name ="Treatment and Sampling Time",
  #                     labels = c("High", "Removal"),
  #                     values = c(24,15))+

  BioR.theme
all_plot_ex
```

```{r}
all_plot_ex <- ggplot() + 
  geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
  geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
  xlab("RDA1 47%") +
  ylab("RDA2 17%") +  
  scale_x_continuous(breaks=seq(-0.3,0.3,0.1), limits=c(-0.3,0.3), sec.axis = dup_axis(labels=NULL, name=NULL),labels = scales::comma) +
  scale_y_continuous(breaks=seq(-0.3,0.3,0.1), limits=c(-0.3,0.3), sec.axis = dup_axis(labels=NULL, name=NULL), labels = scales::comma) +
  
  geom_segment(data=sc_bp_df,
               aes(x=0, y=0, xend=(RDA1*1.5), yend=(RDA2*1.5)),
               colour="black", linewidth=0.3, arrow=arrow()) +
   ggrepel::geom_text_repel(data=sc_bp_df,
                  aes(x=(RDA1*1.3), y=(RDA2*1.3), label=Enviro_var),
                  colour="black") +
  
   # geom_segment(data=species_importance_f,
   #             aes(x=0, y=0, xend=(axis1/2), yend=(axis2/2)),
   #             colour="green4", linewidth=0.3, arrow=arrow())+
   # ggrepel::geom_text_repel(data=species_importance_f,
   #                aes(x=(axis1/2), y=(axis2/2), label=species_importance_f$labels),
   #                colour="black", size=1) +

  # geom_point(data=species_importance_f,
  # aes(x=axis1/2, y=axis2/2),
  # color="black",
  # shape = 23,
  # fill = "green4",
  # size = 3) +
  # ggrepel::geom_text_repel(data = species_importance_f,
  #                 aes(x=axis1/2, y=axis2/2, label = species_importance_f$labels),
  #                colour="black") +
  # 
  # geom_point(data=sc_si_df_1,
  #            aes(x=RDA1, y=RDA2,
  #                colour=sc_si_df_1$NMDSexW.inter2,
  #                fill=sc_si_df_1$NMDSexW.inter2,
  #                shape=sc_si_df_1$NMDSexW.inter2),
  #                size = 2) +
  # 
  # scale_colour_manual(name= "Treatment and Sampling Time",
  #                     labels=c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3"),
  #                     values=c("black",  "black", "black",  "black", "black",  "black", "black",  "black")) +
  # scale_fill_manual(name= "Treatment and Sampling Time",
  #                     labels=c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3"),
  #                     values=c("hotpink4","magenta2","orchid1", "deeppink2","hotpink4","gray47", "gainsboro","gray96")) +
  # scale_shape_manual(name ="Treatment and Sampling Time",
  #                     labels = c("High B1", "High A1", "High A2", "High A3","Removal B1", "Removal A1", "Removal A2", "Removal A3"),
  #                     values = c(24,24,24,24,15,0,1,2))+

  stat_ellipse(data = sc_si_df_1, aes(x=RDA1, y= RDA2, fill = Treatment, color = Treatment), geom="polygon",level=0.95, alpha=0.2) +
  scale_colour_manual(name= "Treatment and Sampling Time",
                      labels=c("High","Removal"),
                      values=c("black",  "black")) +
  scale_fill_manual(name= "Treatment and Sampling Time",
                      labels=c("High", "Removal"),
                      values=c("hotpink4","gray47")) +
  scale_shape_manual(name ="Treatment and Sampling Time",
                      labels = c("High", "Removal"),
                      values = c(24,15))+
  BioR.theme
all_plot_ex
```
